Conversations in time: interactive visualization to explore structured temporal data

### 2021-07-18 tsibbletalk.tex
done with revision
tweaks for publication

% !TeX root = RJwrapper.tex
\title{Conversations in time: interactive visualisation to explore
structured temporal data}
\author{by Earo Wang and Dianne Cook}

@@

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. Nesting of categorical variables produces a hierarchical
structure. The \CRANpkg{tsibbletalk} package is developed to allow a
user to interactively explore temporal data, relative to the nested or
crossed structures. It can help to discover differences between category
levels, and uncover interesting periodic or aperiodic slices. The
package implements a shared \texttt{tsibble} object that allows for
linked brushing between coordinated views, and a shiny module that aids
in wrapping time lines for seasonal patterns. The tools are demonstrated
using two data examples: domestic tourism in Australia and pedestrian
traffic in Melbourne.
}

@@

Temporal data typically arrives as a set of many observational units
measured over time. Some variables may be categorical, containing a
hierarchy in the collection process, that may be measurements taken in
different geographic regions, or types of products sold by the one
company. Exploring these multiple features can be daunting. Ensemble
graphics \citep{unwin_ensemble_2018} bundle multiple views of a data set
together into one composite figure. These provide an effective approach
for exploring and digesting many different aspects of temporal data.
Adding interactivity to the ensemble can greatly enhance the exploration
process.

@@

Systems without the web technology include \CRANpkg{grDevices},
\CRANpkg{loon} \citep{R-loon}, based on Tcl/Tk, and \pkg{cranvas}
\citep{xie_reactive_2014} based on Qt. They offer a wide array of
pre-defined interactions, such as selecting and zooming, to manipulate
plots via mouse action, keyboard strokes, and menus. The
\pkg{cranvastime} package \citep{cheng_enabling_2016} is an add-on to
\pkg{cranvas}, which provides specialised interactions for temporal
data, such as wrapping and mirroring.

The techniques implemented in work described in this paper utilise web
technology, including \CRANpkg{crosstalk}, \CRANpkg{plotly}, and R
\CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance
built on a \texttt{tsibble}. This allows for seamless communication
between different plots of temporal data. The
\texttt{as\_shared\_tsibble()} function turns a \texttt{tsibble} into a
shared instance, \texttt{SharedTsibbleData}, which is a subclass of
\texttt{SharedData} from \CRANpkg{crosstalk}. This is the object driving
data transmission across multiple views. The \CRANpkg{tsibbletalk}
package aims to streamline interactive exploration of temporal data,
with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical variable linking where selecting one or more observations in
one category will broadcast to all other observations in this category.
That is, linking is by key variables: within the time series plot, click
on any data point, and whole line will be highlighted in response. The
\texttt{as\_shared\_tsibble()} uses \texttt{tsibble}'s \texttt{key}
variables to achieve these types of linking.

@@

To illustrate nesting and crossing we use the \texttt{tourism\_monthly}
dataset \citep{tourism} packaged in \CRANpkg{tsibbletalk}. It contains
monthly domestic overnight trips across Australia. The \texttt{key} is
comprised of three identifying variables: \texttt{State},
\texttt{Region}, and \texttt{Purpose} (of trip), in particular
\texttt{State} nesting of \texttt{Region}, together crossed with
\texttt{Purpose}. This specification can be translated as follows:

\begin{Schunk}
\begin{Sinput}
library(tsibbletalk)
tourism_shared <- tourism_monthly %>% 
  as_shared_tsibble(spec = (State / Region) * Purpose)
\end{Sinput}
\end{Schunk}

@@

\noindent There is a three-level hierarchy: the root node is implicitly
Australia, and geographically disaggregated to states and lower-level
tourism regions. A new handy function \texttt{plotly\_key\_tree()} has
been implemented to help explore the hierarchy. It interprets
hierarchies in the shared tsibble's \texttt{spec} as a tree view, built
with \CRANpkg{plotly}. The following code line produces the linked tree
diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual
for the tree hierarchy detangles a group of related series and provides
a bird's eye view of the data organisation.

@@

These line plots are heavily overplotted. To tease apart structure in
the multiple time series, the \texttt{features()} function computes
interesting characteristics, including the measures of trend and
seasonality. These are displayed in the scatterplot at bottom right,
where one dot represents one series.

@@

There is one final step, to compose the three plots into an ensemble of
coordinated views for exploration, shown in Figure
\ref{fig:tourism-linking-fig}. (This is the interactive realisation of
Figure \ref{fig:highlight-retail}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A branch of the tree corresponding to Western Australia was first
  selected. (The names of the regions are a little odd, which is a quirk
  of the data set, but all four areas, Australia's South West, \ldots.,
  correspond to tourist destinations in Western Australia. Hovering over
  the node on the branch brings up the state name.) This generated the
  response in the line plots and the scatterplot that coloured
  corresponding time series and points as blue.
\item
  To enable persistent selection, in oder to compare regions or states,
  ``Shift'' and click on the tree was done, after switching the colour
  to red. This generated the response that points and time series
  corresponding to Sydney were highlighted in red.
\item
  Hovering over the points brings up the label for Sydney.
\end{enumerate}

The domestic tourism sees Sydney as one of the most popular destinations
in realm of business and friends visiting over the years. Despite the
relatively weaker performance in Western Australia, Australia's North
West region sees a strongest upward trend in Business, bypassing Sydney
in some years.

In summary, shared tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} function, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has
cyclical structure, that may correspond to seasonal patterns to be
discovered. The \texttt{index} component of the (shared) tsibble data
forms the basis for exploring seasonality. To investigate for periodic
or aperiodic patterns, series should be wrapped on themselves, where the
index is broken into temporal components like quarter or day. We shall
explore this with pedestrian traffic in Melbourne, Australia.

@@

The city of Melbourne has sensors installed at various locations, to
record hourly counts of pedestrians, in order to capture daily rhythms
of the downtown \citep{ped}. Figure \ref{fig:wrap-ped} shows the first
five months of 2020 foot traffic at four different locations, for three
different time slices, daily, weekly and full five months. Plot
\ref{fig:wrap-ped-1} shows hourly counts from January to May on an
absolute time line, facetted by locations. The stage 3 COVID-19
lockdown, on March 16, is marked by a change of colour. (The
pre-lockdown period is coloured with dark green and lockdown with
orange.) We can see a significant decline in foot traffic at all four
locations. QV Market is less affected probably because this is a major
produce market, an essential service that continued to operate. Bourke
St, a major shopping center, sees a gradual uptick in the last weeks of
the period indicating that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and
wrapping of the series into daily and weekly sections, respectively.
Multiple seasonalities pop out. There tends to be a daily pattern,
especially visible at the main train station, Southern Cross Station.
There is also a weekday vs weekend pattern, also most visible at
Southern Cross Station. These seasonal patterns are still present during
the lockdown, but the magnitude is greatly reduced. Numbers are also
down at the produce market and the shopping center. Birrarung Marr is
the most affected. This is the location of special events, and it is
clear that these have completely disappeared during lockdown.

@@

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
reusable component that could be embedded in any shiny application. In
general, a shiny module provides a vehicle for modularising shiny
applications, relevant for both users and developers. As with all shiny
modules, the first argument in both functions in \CRANpkg{tsibbletalk}
requires a user-supplied id string that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is the server function
\texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, so that the time \texttt{index}
can be retrieved for dissection. The \texttt{period} option semantically
takes a desired number of seasonal periods to be shifted, for example
data shifted by ``1 day'', ``2 days'', or ``1 week'', etc. In other
words, the \texttt{period} defines the grind level. For date-times
(represented by \texttt{POSIXt}), the granularity ranges from fine
``day'' to a much coarser ``year''. The following code snippet generates
Figure \ref{fig:wrap-ped}.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the
slider incremented by 1-day units. The ``play'' button near the end of
slider can automatically animate the slicing and dicing process, walking
the viewer through all 24 hours of the 152 days. Alternatively, users
can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object,
available through the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviours, with nimble user
experience.

@@

The \texttt{as\_shared\_tsibble()} turns a \texttt{tsibble} object to a
shared data class, with an option to express any nesting and crossing
structures from the \texttt{key} attribute. If nesting is found in the
data, the \texttt{plotly\_key\_tree()} creates an interactive
hierarchical tree to help with data overview. This sets the stage for
hierarchical and categorical linking between multiple views from one
shared tsibble.

@@

\address{%
Earo Wang\\
The University of Auckland\\%
Department of Statistics\\
%
%
%
\\\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\\\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}
}

>>>>

% !TeX root = RJwrapper.tex
\title{Conversations in time: interactive visualization to explore
structured temporal data}
\author{by Earo Wang and Dianne Cook}

@@

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. The nesting of categorical variables produces a
hierarchical structure. The \CRANpkg{tsibbletalk} package is developed
to allow a user to interactively explore temporal data, relative to the
nested or crossed structures. It can help to discover differences
between category levels, and uncover interesting periodic or aperiodic
slices. The package implements a shared \texttt{tsibble} object that
allows for linked brushing between coordinated views, and a shiny module
that aids in wrapping timelines for seasonal patterns. The tools are
demonstrated using two data examples: domestic tourism in Australia and
pedestrian traffic in Melbourne.
}

@@

Temporal data typically arrives as a set of many observational units
measured over time. Some variables may be categorical, containing a
hierarchy in the collection process, that may be measurements taken in
different geographic regions, or types of products sold by one company.
Exploring these multiple features can be daunting. Ensemble graphics
\citep{unwin_ensemble_2018} bundle multiple views of a data set together
into one composite figure. These provide an effective approach for
exploring and digesting many different aspects of temporal data. Adding
interactivity to the ensemble can greatly enhance the exploration
process.

@@

Systems without the web technology include \CRANpkg{grDevices},
\CRANpkg{loon} \citep{R-loon}, based on Tcl/Tk, and \pkg{cranvas}
\citep{xie_reactive_2014} based on Qt. They offer a wide array of
pre-defined interactions, such as selecting and zooming, to manipulate
plots via mouse action, keyboard strokes, and menus. The
\pkg{cranvastime} package \citep{cheng_enabling_2016} is an add-on to
\pkg{cranvas}, which provides specialized interactions for temporal
data, such as wrapping and mirroring.

The techniques implemented in the work described in this paper utilize
web technology, including \CRANpkg{crosstalk}, \CRANpkg{plotly}, and R
\CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance
built on a \texttt{tsibble}. This allows for seamless communication
between different plots of temporal data. The
\texttt{as\_shared\_tsibble()} function turns a \texttt{tsibble} into a
shared instance, \texttt{SharedTsibbleData}, which is a subclass of
\texttt{SharedData} from \CRANpkg{crosstalk}. This is an R6 object
driving data transmission across multiple views, due to its mutable and
lightweight properties. The \CRANpkg{tsibbletalk} package aims to
streamline interactive exploration of temporal data, with the focus of
temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical variable linking, where selecting one or more observations
in one category will broadcast to all other observations in this
category. That is, linking is by key variables: within the time series
plot, click on any data point, and the whole line will be highlighted in
response. The \texttt{as\_shared\_tsibble()} uses \texttt{tsibble}'s
\texttt{key} variables to achieve these types of linking.

@@

To illustrate nesting and crossing we use the \texttt{tourism\_monthly}
dataset \citep{tourism} packaged in \CRANpkg{tsibbletalk}. It contains
monthly domestic overnight trips across Australia. The \texttt{key} is
comprised of three identifying variables: \texttt{State},
\texttt{Region}, and \texttt{Purpose} (of the trip), in particular
\texttt{State} nesting of \texttt{Region}, crossed together with
\texttt{Purpose}. This specification can be translated as follows:

\begin{Schunk}
\begin{Sinput}
library(tsibble)
library(tsibbletalk)
tourism_shared <- tourism_monthly %>% 
  as_shared_tsibble(spec = (State / Region) * Purpose)
\end{Sinput}
\end{Schunk}

@@

\noindent There is a three-level hierarchy: the root node is implicitly
Australia, geographically disaggregated to states, and lower-level
tourism regions. A new handy function \texttt{plotly\_key\_tree()} has
been implemented to help explore the hierarchy. It interprets
hierarchies in the shared tsibble's \texttt{spec} as a tree view, built
with \CRANpkg{plotly}. The following code line produces the linked tree
diagram (left panel of Figure \ref{fig:tourism-linking-fig}). The visual
for the tree hierarchy detangles a group of related series and provides
a bird's eye view of the data organization.

@@

These line plots are heavily overplotted. To tease apart structure in
the multiple time series, the \texttt{features()} function computes
interesting characteristics, including the measures of trend and
seasonality. These are displayed in the scatterplot at the bottom right,
where one dot represents one series.

@@

There is one final step, to compose the three plots into an ensemble of
coordinated views for exploration, shown in Figure
\ref{fig:tourism-linking-fig}. (This is the interactive realization of
Figure \ref{fig:highlight-retail}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A branch of the tree corresponding to Western Australia was first
  selected. (The names of the regions are a little odd, which is a quirk
  of the data set, but all four areas, Australia's South West, \ldots.,
  correspond to tourist destinations in Western Australia. Hovering over
  the node on the branch brings up the state name.) This generated the
  response in the line plots and the scatterplot that colored
  corresponding time series and points as blue.
\item
  To enable persistent selection, in oder to compare regions or states,
  ``Shift'' and click on the tree was done, after switching the color to
  red. This generated the response that points and time series
  corresponding to Sydney were highlighted in red.
\item
  Hovering over the points brings up the label for Sydney.
\end{enumerate}

Domestic tourism sees Sydney as one of the most popular destinations in
the realm of business and friends visiting over the years. Despite the
relatively weaker performance in Western Australia, Australia's North
West region sees a strongest upward trend in business, bypassing Sydney
in some years.

In summary, shared tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for the effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} function, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has a
cyclical structure, that may correspond to seasonal patterns to be
discovered. The \texttt{index} component of the (shared) tsibble data
forms the basis for exploring seasonality. To investigate for periodic
or aperiodic patterns, series should be wrapped on themselves, where the
index is broken into temporal components like quarter or day. We shall
explore this with pedestrian traffic in Melbourne, Australia.

@@

The city of Melbourne has sensors installed at various locations, to
record hourly counts of pedestrians, in order to capture the daily
rhythms of the downtown \citep{ped}. Figure \ref{fig:wrap-ped} shows the
first five months of 2020 foot traffic at four different locations, for
three different time slices, daily, weekly and full five months. Plot
\ref{fig:wrap-ped-1} shows hourly counts from January to May on an
absolute timeline, facetted by locations. The stage 3 COVID-19 lockdown,
on March 16, is marked by a change of color. (The pre-lockdown period is
colored with dark green and lockdown with orange.) We can see a
significant decline in foot traffic at all four locations. QV Market is
less affected probably because this is a major produce market, an
essential service that continued to operate. Bourke St, a major shopping
center, sees a gradual uptick in the last weeks of the period indicating
that people were getting back into the shops.

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show the slicing
and wrapping of the series into daily and weekly sections, respectively.
Multiple seasonalities pop out. There tends to be a daily pattern,
especially visible at the main train station, Southern Cross Station.
There is also a weekday vs weekend pattern, also most visible at
Southern Cross Station. These seasonal patterns are still present during
the lockdown, but the magnitude is greatly reduced. Numbers are also
down at the produce market and the shopping center. Birrarung Marr is
the most affected. This is the location of special events, and it is
clear that these have completely disappeared during the lockdown.

@@

This shiny module, decoupled to \texttt{tsibbleWrapUI()} and
\texttt{tsibbleWrapServer()}, presents a clean interface and forms a
reusable component that could be embedded in any shiny application. In
general, a shiny module provides a vehicle for modularising shiny
applications, relevant for both users and developers. As with all shiny
modules, the first argument in both functions in \CRANpkg{tsibbletalk}
requires a user-supplied id string that must be unique. The UI function
\texttt{tsibbleWrapUI()} simply shows a slider that animates or controls
the number of periods to be diced. The workhorse is the server function
\texttt{tsibbleWrapServer()}, encapsulating the algorithm that
transforms data and sends messages to update the plot accordingly. The
\texttt{plot} argument expects a \texttt{ggplot} or \texttt{plotly}
object, where one can plot data using either lines or other graphical
elements (such as boxplots). As the function name suggests, a (shared)
tsibble is needed to start the engine, so that the time \texttt{index}
can be retrieved for dissection. The \texttt{period} option semantically
takes a desired number of seasonal periods to be shifted, for example
data shifted by ``1 day'', ``2 days'', or ``1 week'', etc. In other
words, the \texttt{period} defines the grind level. For date-times
(represented by \texttt{POSIXt}), the granularity ranges from fine
``day'' to a much coarser ``year''. The following code snippet generates
Figure \ref{fig:wrap-ped}. The creation of the \texttt{pedestrian20}
data is available in supplementary R files.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the
slider incremented by 1-day units. The ``play'' button near the end of
the slider can automatically animate the slicing and dicing process,
walking the viewer through all 24 hours of the 152 days. Alternatively,
users can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object,
available through the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviors, with a nimble
user experience.

@@

The \texttt{as\_shared\_tsibble()} turns a \texttt{tsibble} object to a
shared data class, with an option to express any nesting and crossing
structures from the \texttt{key} attribute. If nesting is found in the
data, the \texttt{plotly\_key\_tree()} creates an interactive
hierarchical tree to help with the data overview. This sets the stage
for hierarchical and categorical linking between multiple views from one
shared tsibble.

@@

\address{%
Earo Wang\\
The University of Auckland\\%
Department of Statistics\\
%
%
%
\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}%
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}
